### 2014-01-27 paper/chemokine_main.tex
Plos ONE Submission

@@

% Title must be 150 characters or less
\begin{flushleft}
{\Large
\textbf{Spatially explicit model of the lymphocyte diaspora in influenza-infected lung reveals thresholds on chemokine directed migration}
}
% Insert Author names, affiliations and corresponding author email.
\\
Drew Levin$^{1,\ast}$, 
Stephanie Forrest$^{1}$, 
Soumya Banerjee$^{1}$,
Candice Clay$^{2}$, 
Melanie Moses$^{1}$, 
Frederick Koster$^{1,2}$
\\
\bf{1} Department of Computer Science, University of New Mexico, Albuquerque, NM, USA
\\
\bf{2} Lovelace Respiratory Research Institute, Albuquerque, NM, USA
\\
$\ast$ E-mail: Corresponding drew@cs.unm.edu
\end{flushleft}

@@

During the primary immune response, clearance of influenza in the lung requires the homing of activated CD8 T cells from regional lymph nodes to infected foci comprising a fraction of the total lung.  T cell navigation to infected foci is undirected but emigration into tissue is directed by local cytokine and chemokine production from infected epithelial cells.  We focused on the efficiency of local chemokines to induce migration of activated CD8 T cells using a spatially explicit agent-based model (ABM) without constructing a model containing all elements of the immune response.

To determine local chemokine production, avian H5N1, seasonal H1N1, and 2009 pandemic influenza strains were used \textit{in vitro} to induce the secretion of CXCL10 (IP-10) and CCL5 (RANTES) in human bronchial epithelial cells.  A differential equation model was fit to empirical chemokine production rates and coupled with published T cell parameters to calibrate the spatial model to test inter-strain variation on T cell recruitment in the lungs.  The modeled immune response is unable to clear the pandemic strain due to its high rate of viral production.

The spatial nature of the ABM model reveals unique challenges to T cell recruitment not appreciated in standard differential equation models.  Infected cells can become isolated in expanding plaques, impeding T cell search, even though the T cells could directionally migrate to low levels of chemokine.  A key limitation imposed on T cells is illustrated by their failure to clear the pandemic H1N1 virus after day 6, when T cells became inefficient in large infected foci. This spatially explicit model is consistent with efficient T cell recruitment to small infected foci in the lung, but as foci become large T cell search becomes inefficient, thus emphasizing the importance of enhanced control early in infection.

@@

Clearance of influenza from the lung depends strongly on the efficiency with which T cells travel from the lymph node to the site of infection.  However, T cells in the lung must navigate the large branching bronchial network.  This maze-like structure is a challenge for T cell search, and the vast majority of T cells exit the lung without encountering infection.  Infected cells produce cytokines and chemokines, which signal T cells to exit the vasculature and enter tissue.  We developed an agent-based model to quantify how these molecular signals affect search.  Our experimental data show that two chemokines, IP-10 and RANTES, are important for T cell search.  We incorporated the data in our spatially explicit model, which allowed us to identify challenges to the search process not obvious in conventional differential equation models.  We ran three sets of simulations, using parameters from seasonal influenza, avian influenza, and 2009 pandemic influenza.  The simulated immune response contains the first two infections but is unable to contain the highly virulent pandemic strain due to the rapidly expanding plaque size.  Direct interventions to improve this T cell search process are unlikely to be effective, and therefore vaccination is required to contain more virulent strains.

@@

The adaptive immune response induced during acute influenza A infection is a complex combination of defense mechanisms.  Understanding the behavior and interactions of each cellular component may lead to improved vaccines and strategies to control immunopathology.   The innate immune response is critical for early control of viral replication, followed by antibody-mediated viral neutralization \cite{Kohlmeier2009a, Joo2008, Oslund2011}.   Complete resolution of influenza pneumonia depends on antigen-specific CD8+ T effector cells \cite{Cerwenka1999, Kim2011}.  The murine model has been critical in the reductive analysis of each phase and component, permitting temporally explicit descriptions of the induction phase in the regional lymph node \cite{Miller2003, Allan2006, Ingulli2009} where recruitment of clonal precursor CD8 T cells is highly efficient \cite{VanHeijst2009}.   On approximately day 5 after infection, activated CD8 T cells are released from the secondary lymphoid tissue \cite{Miao2010a} and distributed throughout the body in a process known as the ‘lymphocyte diaspora’ \cite{Marshall2001}.

The process of recruitment of activated CD8 effector cells to infected sites in the lung is not completely understood.  Initial localization and extravasation may depend on a number of inflammatory signals but viral antigen does not appear to induce recruitment into tissue \cite{Topham2001}.   Effector cells must localize to sites of viral replication [4], but it is not clear whether cells passing through uninfected lung tissue leave through the pulmonary vein or exit the capillary bed and leave the lung through lymphatic channels. Chemotactic proteins play a major role in recruitment in both acute infections and chronic inflammatory diseases \cite{Bromley2008, Medoff2005} where recruited T cells mediate either protection or immunopathology.  Infected epithelial cells secrete chemokines \cite{Chan2005} and contact with CD8 T cells also triggers secretion \cite{Zhao2000}. Activated CD8 T effector cells upregulate chemotactic protein receptors, particularly CXCR3 and CCR5 \cite{Hoji2005, Groom2011a}, responding to their primary ligands CXCL10 (IP-10) and CCL5 (RANTES) respectively.

@@

We ask how small foci of infected tissue, scattered through a very large space of uninfected tissue, can attract and retain limited numbers of activated CD8 T cells.  Our model does not consider important features of tissue immunity such as T cell proliferation, resident T cells, tertiary lymphoid structures, and regulatory interactions with other cell phenotypes \cite{McGill2009}.  The model does not aim to predict control of viral replication but only to compare simulated outcomes among three influenza strains arising from the efficacy of the chemotactic signals.  We use chemotactic protein data from in vitro cultures of bronchial epithelial cells infected with three strains of influenza virus with different replication rates \cite{Mitchell2011} previously described in our laboratory.  Using known parameters from the extensive literature on the murine model, we offer several insights on the control of viral replication in lung tissue derived from spatially explicit simulations.

@@

\subsection*{Delay Differential Equation Model}

We estimate chemokine production rates by adapting the delay differential equation model of influenza infection described in \cite{Mitchell2011} by adding one new equation to model chemokine production.  Initial population sizes and parameter values are taken from the previous study.
{\footnotesize
\begin{equation}
\begin{aligned}
\dot{T} &= - \beta T V \\
\dot{I_1} &= \beta T V - \beta T[t-\tau_1]V[t-\tau_1] \\
\dot{I_2} &= \beta T[t-\tau_1]V[t-\tau_1] - \delta I_2 \\
\dot{V} &= \frac{p}{1+eF} I_2  - \beta T V  \\
\dot{F} &=  I_1[t-\tau_2] \\
\dot{C} &= r I[t-\tau_3] - d C \\
\end{aligned}
\label{eq:dde}
\end{equation}
\vspace{.05in}
}

Table \ref{tab:dde} summarizes population and parameter values and descriptions.  Strain-specific values for $r$ were found by fitting the equations to experimental data (see Results and Table \ref{tab:strains}).

@@

Computational modeling used CyCells \cite{Warrender2006}, a modeling platform for two- or three-dimensional agent-based simulations of the immune response. A simplified model of CD8 T cell activation and recirculation (Fig.~\ref{fig:systemchart}) was implemented in CyCells (Fig.~\ref{fig:modelchart}), and simulations measured efficiency of infection clearance under different environmental conditions. The lung was represented as a two-dimensional sheet of healthy epithelial cells. Vasculature was represented as a binary tree with fourteen levels of bifurcations with the origin at the bronchial lymph node complex. Activated T cells descend through the vascular tree until cytokine signal is detected on the local endothelium, at which point they exit the vasculature and follow the chemotactic gradient to the site of infection. T cells that do not encounter cytokine recirculate to the lymph node.  When a T cell contacts an infected epithelial cell at the site of infection it induces apoptosis.

The simulation begins when a single cell in the center of the tissue is infected. After the eclipse phase (incubation), the infected cell begins secreting virus and chemokine according to the virus strain and chemokine type. Virus diffuses locally, infecting nearby cells, and continuing the cycle. Chemokine diffuses from secreting cells, creating a circle of stimulation around the infected region. After a five day delay to simulate lymph node stimulation and T cell proliferation, activated T cells exit the lymph node and circulate through the vasculature to the tissue. Because T  cells cannot choose their path through the branching network, we assume they arrive in the capillary bed at random locations.

@@

In the model, epithelial cells are stationary and can be in one of five different states: \emph{healthy}, \emph{virus-incubating}, \emph{virus-expressing}, \emph{apoptotic}, and \emph{dead}. \emph{Healthy} cells remain unchanged unless infected by virus. Once infected, the cell transitions from {incubating} to \emph{expressing}. \emph{Expressing} cells secrete virus and chemokine for approximately 17 hours and then die. \emph{Expressing} cells become \emph{apoptotic} if they encounter activated T cells. Apoptoic cells continue to secrete virus until they die one hour after their transition. \emph{Dead} cells remain inert and do not regenerate over the course of an infection.

T cells have three states. \emph{Circulating} T cells begin to emerge from the lymph node at five days post infection. \textit{Emigrating} T cells arrive at a random location on the lung's surface, wander randomly in the tissue for 10 minutes, and transition to \emph{circulating} in the absence of chemokine. \emph{Circulating} cells spend six minutes recirculating to the lymph node, transition to \emph{emigrating} and are reintroduced to a new random location in the lung. When a \emph{emigrating} T cell encounters chemokine, it converts to \emph{chemotaxing} and begins following the chemotactic gradient to the source of infection. \emph{Chemotaxing} T cells move continuously up the gradient, inducing apoptosis if they encounter \emph{expressing} epithelial cells. \emph{Chemotaxing} cells decay exponentially with an average lifespan of two hours.

The model contains two kinds of particles: virus and chemokine. Both are produced at constant rates by \emph{expressing} epithelial cells.  Virus diffuses through the lung tissue, infecting healthy cells probabilistically according to the local virus concentration. Chemokine diffuses across the tissue but has no direct effect beyond activating T cells. Both particle types decay exponentially.  IgM is modeled by increasing the viral decay rate by a factor of three after the third day.

Parameters that are consistent between every model are shown in Table \ref{tab:parameters}.  Strain-specific values are shown in Table \ref{tab:strains}.

@@

Chemokine secretion:  Epithelial cell culture and supernatant collection was performed as described \cite{Mitchell2011}.  Briefly, undifferentiated human tracheal epithelial cells (University of Miami) were cultured for 4 weeks to achieve fully differentiated confluent monolayers on collagen-coated transwell inserts, or commercial differentiated human bronchial epithelial cells (EpiAirway Tissue, MatTek Corp., Ashland, MA) used immediately upon receipt, were infected at an MOI of 0.01 with either seasonal H1N1 virus A/New Caledonia/20/99 (sH1N1), the 2009 H1N1 pandemic strain A/California/04/09 (pH1N1), or avian H5N1 virus A/Hong Kong/483/97 (aH5N1) derived from a fatal human infection.  Basal media was collected from previously undisturbed triplicate or quadruplicate wells at 0, 6, 10, 12, 16, 20, 24, 30, 36, 42, 48, and 72 hours after infection, and stored at -80C until assay.  Subsequently, apical fluid for virus secretion was collected before and after treatment of the monolayer with protease (Pronase, Sigma) to optimize the collection of infectious virus \cite{Mitchell2011}.  Quantitative viral culture was performed by standard plaque assay.  Quantitative chemokine levels were performed in 30 µl aliquots for a panel of chemokines (IL-8, MCP-1, MIP-1$\alpha$, MIP-1$\beta$, RANTES, IP-10, eotaxin) and cytokines (interferon-gamma, IL-1$\alpha$, IL-1$\beta$, IL-1RA, IL-2, IL-3, IL-4, IL-6, IL-10, IL-12p40, IL-15, IL-17, TNF$\alpha$) (Luminex Assay®, Luminex Corp.) and reported as ng/mL basal media sampled from a total volume of 4 mL.  Only IP-10, RANTES, and TNF$\alpha$ showed increases in production.  TNF was ignored as its effects are outside the scope of this paper.

@@

To provide estimates of chemokine concentrations and secretion rates present in lung tissue, chemokine levels were measured at 4-6h intervals during the first 48 h of infection in wells containing approximately one million human bronchial epithelial cells (Fig.~\ref{fig:data}).  The dynamic viral loads at these intervals in these cultures infected with seasonal H1N1 virus, pandemic H1N1 virus, and avian H5N1 virus have been reported previously \cite{Mitchell2011}.  IP-10 concentration increases were observed by 8h post-infection (p.i.), and RANTES by 16h p.i..  To estimate per-cell production rates, we extended the ordinary differential equation (ODE) model of Ref. \cite{Mitchell2011} to represent chemokine production from infected cells (Eq.~\ref{eq:dde}).  Model fits (Table \ref{tab:strains}) were computed for three strains (Fig.~\ref{fig:data}) using Matlab's \texttt{nlinfit} function (Levenberg-Marquardt algorithm).  The resulting chemokine production values were used in the CyCells ABM.  Best-fit expression rates were similar for all strains except for significantly higher RANTES production in aH5N1.  There is no positive correlation between viral production and induced chemokine production across the three strains.

\subsection*{T cell sensitivity to chemokine}

The model simulates a chemokine gradient surrounding the infected focus (Fig.~\ref{fig:cycells}), based on the calculated per-infected cell secretion rate (Table~\ref{tab:strains}) and known chemical parameters for a 10 kDa protein (Table~\ref{tab:parameters}).  T cell sensitivity depends on receptor density \cite{Desmetz2006} and this was assumed to be constant.

Because this parameter is unknown, we simulated T cell sensitivity levels ranging over 10 orders of magnitude and found a threshold (Fig.~\ref{fig:sensitivity}) at a concentration of 100 $ng/ml$ (10 $ng/ml$ in aH5N1), beyond which there is no detectable effect on model behavior (model variance is discussed in Text S1 2.1.).  We then set the sensitivity to 10 $ng/mL$ for all future runs (1 $nM$ concentration assuming a chemokine molecular weight of 10 $kDa$) \cite{Gao2003}.

% Because more than one T cell at each infected epithelial cell would not increasing effective target cell death, there may be a threshold of immigrant T cells above which no additional control of viral replication is possible.

@@

Spatial effects of viral and chemokine diffusion play an important role in both the spreading and clearing of infections.  Free virus particles diffuse from virus secreting cells and infect healthy cells.  Chemokine produced by infected cells attracts T cells to the infected cells.  Although virus is produced at a higher rate than chemokine, its larger size diffuses much more slowly, while chemokine decays more quickly.  These countervaling effects result in similar spatial profiles for the two particle types (Fig.~\ref{fig:cycells}). Until day 4 the plaque is dominated by active (incubating and secreting) cells, whereas dead cells are rare. Over time, cells in the plaque's interior die, and active cells form a decreasing proportion of the plaque. T cells arrive at day 5 and begin killing the virus-secreting cells. By day 6 many expressing cells have been eliminated and the plaque is dominated by dead cells.  Refer to the Videos S1-S3 to see a representative simulation of seasonal influenza.

In aH5N1, the plaque is dense, allowing T cells to find secreting cells easily, and infection is eliminated.  However, secreting cells were not eliminated in either H1N1 simulations (Fig. S1).  Secreting cells accounted for at most 10\% of the active cell population and less than 1\% of the total plaque at 6 days p.i..  T cells still accumulate, but they arrive more slowly than the plaque is growing, leading to lower average T cell killing rate.  Further, the regions of concentrated chemokine lag behind the cell and virus spatial layout.  It takes time for newly secreting cells to produce chemokine while pockets of high chemokine density are slow to decay.  Thus, T cells can fail to detect cellular changes in the plaque.  Taken together, the delayed response and low proportion of virus secreting cells prevent T cells from completely clearing the infection.  Furthermore, increasing chemokine production fails to control the runaway pH1N1 infection (data not shown) suggesting that the effect of high viral production rates dominate low chemokine concentrations in this scenario.

T cells are unable to kill cells that have not yet presented antigen.  At day 5.5, the ratio of secreting cells to the total plaque size is high (Fig.~\ref{fig:cycells}-\ref{fig:plaquesize}). By day 7, this ratio is very small for both the pH1N1 and sH1N1 strains, making it much more difficult for searching T cells to find the secreting cells.  The high replication rate of pH1N1 enhances this effect (Fig.~\ref{fig:plaquesize}C) and the T cells can control (but not eliminate) the sH1N1 infection.  Video S4 illustrates this event.

A cell infected with pH1N1 produces new virus at the rate of 5.08e-3 PFU/s \cite{Mitchell2011}.  That is, in each secreting cell a new viral particle is produced approximately every 200 seconds.  Assuming that secretion continues for one hour after apoptosis is initiated, the best a T cell could do is limit production to 18 new viral particles.  Thus, T cells alone cannot contain the pH1N1 infection.  In contrast, a sH1N1 virus-secreting cell produces a new virus particle every 2,632 seconds, allowing T cells to limit a single infected cell to 1.4 viral particles in the one-hour window.  Avian H5N1 virus-secreting cells produce only 0.2 viral particles in the interval after induced apoptosis.

>>>>

\usepackage{color}
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage{ulem}

\definecolor{dkred}{rgb}{0.75,0,0}
\definecolor{dkgreen}{rgb}{0,0.5,0}
\definecolor{dkblue}{rgb}{0,0,0.75}
\definecolor{dkpurple}{rgb}{.375,0,.375}
\definecolor{gray}{rgb}{0.5,0.5,0.5}

\newcommand{\removed}[1]{{\color{dkred}\sout{#1}}}
\newcommand{\drew}[1]{{\color{dkgreen}#1}}
\newcommand{\fred}[1]{{\color{dkblue}#1}}
\newcommand{\steph}[1]{{\color{dkpurple}#1}}

@@

% Title must be 150 characters or less
\begin{flushleft}
{\Large
\textbf{Spatially explicit model of the efficiency of the lymphocyte diaspora in influenza-infected lung reveals thresholds on chemokine directed migration}
}
% Insert Author names, affiliations and corresponding author email.
\\
Drew Levin$^{1,\ast}$, 
Stephanie Forrest$^{1}$, 
Soumya Banerjee$^{1}$,
Candice Clay$^{2}$, 
Melanie Moses$^{1}$, 
Frederick Koster$^{1,2}$
\\
\bf{1} Department of Computer Science, University of New Mexico, Albuquerque, NM, USA
\\
\bf{2} Lovelace Respiratory Research Institute, Albuquerque, NM, USA
\\
$\ast$ E-mail: Corresponding drew@cs.unm.edu
\end{flushleft}

@@

During the primary immune response, clearance of influenza in the lung requires the homing of activated CD8 T cells from regional lymph nodes to infected foci comprising a small fraction of the total lung.  T cell navigation towards infected foci is undirected but emigration into tissue is directed by local cytokine and chemokine production from infected epithelial cells.  Weuse a spatially explicit agent-based model (ABM) to examine the efficiency of the recruitment of activated CD8 T cells by local chemokine signalling.

To determine local chemokine production, avian H5N1, seasonal H1N1, and 2009 pandemic influenza strains were used \textit{in vitro} to induce the secretion of CXCL10 (IP-10) and CCL5 (RANTES) in human bronchial epithelial cells.  These data were modeled to estimate chemokine production rates. The estimates were combined in the spatial model to test viral variation of T cell recruitment in the lungs.

The spatial nature of the ABM model reveals unique challenges to T cell recruitment not appreciated in standard differential equation models.  Infected cells can become isolated in expanding plaques, impeding T cell search, even though the T cells could directionally migrate to low levels of chemokine.  This spatially explicit model describes efficient T cell recruitment to small infected foci in the lung, but as foci become large T cell search becomes inefficient, thus emphasizing the importance of enhanced control early in infection. A key limitation imposed on T cells is illustrated by their failure to clear the pandemic H1N1 virus after day 6, when T cells became inefficient in finding infected cells in large foci.  We conclude that viral dynamics, rather than chemokine dynamics, dominate the efficiency of T cell recruitment in determining whether the influenza infection is controlled.

@@

Clearance of influenza from the lung depends critically on the efficiency with which T cells travel from the lymph node to the site of infection.  However, T cells in the lung must navigate the large branching bronchial network.  This maze-like structure is a challenge for T cell search, and the vast majority of T cells exit the lung without encountering infection.  Infected cells produce cytokines and chemokines, which signal T cells to exit the vasculature and enter tissue.  We developed an agent-based model to quantify how these molecular signals affect search.  Our experimental data show that two chemokines, IP-10 and RANTES, are secreted by infected human tracheal epithelial cells.  We incorporated this data in our spatially explicit model, which allowed us to identify challenges to the search process not obvious in conventional differential equation models.  We ran three sets of simulations, using parameters from seasonal influenza, avian influenza, and 2009 pandemic influenza.  The simulated immune response controls and eradicates the first two infections in 10 days but is unable to contain the highly virulent pandemic strain due to the rapidly expanding plaque size.  Interventions are unlikely to improve this chemokine-directed T cell search process in the lung but vaccination against more virulent strains will increase the number of searching cells earlier in the response.

@@

The adaptive immune response induced during acute influenza A infection is an integrated complex of combination of defense mechanisms.  Understanding the behavior and interactions of each cellular component may lead to improved vaccines and strategies to control the infection and immunopathology.   The innate immune response is critical for early control of viral replication, followed by antibody-mediated viral neutralization \cite{Kohlmeier2009a, Joo2008, Oslund2011}.   Complete resolution of influenza pneumonia depends on antigen-specific CD8+ T effector cells \cite{Cerwenka1999, Kim2011}.  The murine model has been critical in the reductive analysis of each phase and component, permitting temporally explicit descriptions of the induction phase in the regional lymph node \cite{Miller2003, Allan2006, Ingulli2009} where recruitment of clonal precursor CD8 T cells is highly efficient \cite{VanHeijst2009}.   On approximately day 5 after infection, activated CD8 T cells are released from the secondary lymphoid tissue \cite{Miao2010a} and distributed throughout the body in a process known as the ‘lymphocyte diaspora’ \cite{Marshall2001}.

CD8 T cells conduct two `searches' in two different tissue environments, first to encounter antigen-loaded dendritic cells in the lymph node, and second to encounter local inflammatory signals in infected lung.  Search problems in the lymph node have been simulated using live cell imaging data to provides reliable parameters of cell movement \cite{Vroomans2012}.  The T precursor cell must encounter on average hundreds of irrelevant cells before it contacts the loaded DC, but the distance traveled is relatively short, and the direction is likely guided by chemokine signals.

The second search, i.e. the process of recruiting activated CD8 effector cells to infected sites in the lung,  is different and not as well understood.  In contrast to the dense lymph node volume, lung infection dynamics occur on the surface of epithelial cells, the size of the tissue layer being comparable to the size of a tennis court.  Initial localization and extravasation may depend on a number of inflammatory signals but viral antigen does not appear to induce recruitment into this tissue \cite{Topham2001}.   Moreover, the T cells leaving the lymph node are released into the very large systemic vascular `space' with no initial guiding signals.  Effector cells must localize to sites of viral replication \cite{Cerwenka1999}, but it is not clear whether cells passing through uninfected lung tissue leave through the pulmonary vein or exit the capillary bed and leave the lung through lymphatic channels. Chemotactic proteins play a major role in recruitment in both acute infections and chronic inflammatory diseases \cite{Bromley2008, Medoff2005, Castellino2006 ,Gunn1998, Okada2005} where recruited T cells mediate either protection or immunopathology.  Infected epithelial cells secrete chemokines \cite{Chan2005} and contact with CD8 T cells also triggers secretion \cite{Zhao2000}. Activated CD8 T effector cells upregulate chemotactic protein receptors, particularly CXCR3 and CCR5 \cite{Hoji2005, Groom2011a}, responding to their primary ligands CXCL10 (IP-10) and CCL5 (RANTES) respectively.

@@

We ask how small foci of infected tissue, scattered through a very large space of uninfected tissue, can attract and retain limited numbers of activated CD8 T cells.  Our model does not consider important features of tissue immunity such as T cell proliferation, resident T cells, and regulatory interactions with other cell phenotypes \cite{McGill2009}.  The model does not aim to predict control of viral replication but only to compare simulated outcomes among three influenza strains arising from the efficacy of the chemotactic signals.

Previously we developed an agent-based model of the lymphocyte diaspora focusing on the production and diffusion of a generic inflammatory signal without experimental data to calibrate the model \cite{Banerjee2011}.  Here we refine that model and incorporate chemotactic protein data from in vitro cultures of human bronchial epithelial cells infected with three strains of influenza virus with different replication rates \cite{Mitchell2011} to focus on two features of searching in the large lung space.  First, functional thresholds of the chemotactic proteins were surprisingly low in our model.  Second there appears to be a threshold size of the infected focus above which the T cell search becomes inefficient.

@@

@@

Computational modeling used CyCells \cite{Warrender2006}, a modeling platform for two- or three-dimensional agent-based simulations of the immune response. A simplified model of CD8 T cell activation and recirculation (Fig.~\ref{fig:systemchart}) was implemented in CyCells (Fig.~\ref{fig:modelchart} and \ref{fig:variance}), and simulations measured efficiency of infection clearance for three different viruses. The lung was represented as a two-dimensional sheet of healthy epithelial cells (S1.4).  Vasculature in a mouse is a binary tree with fourteen levels of bifurcation with the origin at the bronchial lymph node complex.  Because we assume activated T cells descend at random into branches of the vascular network, T cells are introduced uniformly at random across the entire lung surface (Fig.~\ref{fig:systemchart}).  If T cells detect cytokine signal on the local endothelium they remain in tissue and follow the chemotactic gradient to the site of infection. T cells that do not encounter cytokine recirculate to the lymph node.  When a T cell contacts an infected epithelial cell at the site of infection it induces apoptosis.  Unimpeded, the infected secretes virus for 16.7 hours and then dies.

The simulation begins when a single cell in the center of the tissue is infected. After the eclipse phase (incubation), the infected cell begins secreting virus and chemokine according to the virus strain and chemokine type as determined by simulations reported in S1. Virus diffuses locally, infecting nearby cells, and continuing the cycle. Chemokine diffuses from secreting cells according to kinetics described below, creating a circle of stimulation around the infected region. After a five day delay to simulate lymph node stimulation and T cell proliferation, activated T cells begin to exit the lymph node at a steady rate and circulate through the vasculature to the tissue.  Further details of the CyCells modeling environment are described in S1.3.

Computational limitations force us to implement our model on the scale of a mouse.  Parameters affected by this choice include the T cell production rate, T cell circulation time, and the total size of the lung (Table \ref{tab:parameters}).  The surface area of a mouse lung is approximately 100$cm^2$.

@@

In the model (Fig.~\ref{fig:modelchart}), epithelial cells are stationary and are described by one of five sequential states: \textit{healthy}, \textit{virus-incubating}, \textit{virus-expressing}, \textit{apoptotic}, and \textit{dead}. \textit{Healthy} cells remain unchanged unless infected by virus. Once infected, the cell transitions from \textit{incubating} to \textit{expressing} after a 10 hour incubation delay. \textit{Expressing} cells secrete virus and chemokine for a fixed 16.7 hours and then die. \textit{Expressing} cells initiate \textit{apoptosis} sooner if they are contacted by activated T cells. Apoptoic cells continue to secrete virus and chemokine and then die after one hour. \textit{Dead} cells take up space and do not regenerate over the course of an infection.

T cells are described by two states. \textit{Circulating} T cells begin to emerge from the lymph node at five days post infection and arrive at a random location on the lung's surface after six seconds of travel.  If chemokine is not encountered, the T cell remains a circulating cell, returning to a new location in the lung after another six seconds.  If a circulating cell encounters chemokine, it converts to \textit{chemotaxing} and begins following the chemotactic gradient to the infected region. \textit{Circulating} T cells decay exponentially with an average lifespan of three days.  \textit{Chemotaxing} T cells move continuously up the gradient through the two dimensional lung endothelium, inducing apoptosis in the \textit{expressing} epithelial cells. \textit{Chemotaxing} cells decay exponentially with an average lifespan of two hours.

The model contains two kinds of particles: virus and chemokine. Both are produced at constant rates by \textit{expressing} epithelial cells.  Virus diffuses through the lung tissue, infecting healthy cells at a rate according to the local virus concentration. Chemokine diffuses across the tissue but has no direct effect beyond activating T cells. Both particle types decay exponentially.  IgM is modeled by increasing the viral decay rate by a factor of ten after the third day.

Parameters that are consistent between every model are shown in Table \ref{tab:parameters}.  Strain-specific values are shown in Table \ref{tab:strains}.  Further details regarding the model definition are included in S1.4.

@@

Chemokine secretion:  Epithelial cell culture and supernatant collection was performed as described \cite{Mitchell2011}.  Briefly, undifferentiated human tracheal epithelial cells (University of Miami) were cultured for 4 weeks to achieve fully differentiated confluent monolayers on collagen-coated transwell inserts, or commercial differentiated human bronchial epithelial cells (EpiAirway Tissue, MatTek Corp., Ashland, MA) used immediately upon receipt, were infected at an MOI of 0.01 with either seasonal H1N1 virus A/New Caledonia/20/99 (sH1N1), the 2009 H1N1 pandemic strain A/California/04/09 (pH1N1), or avian H5N1 virus A/Hong Kong/483/97 (aH5N1) derived from a fatal human infection.  Basal media was collected from previously undisturbed triplicate or quadruplicate wells at 0, 6, 10, 12, 16, 20, 24, 30, 36, 42, 48, and 72 hours after infection, and stored at -80C until assay.  Subsequently, apical fluid for virus secretion was collected before and after treatment of the monolayer with protease (Pronase, Sigma) to optimize the collection of infectious virus \cite{Mitchell2011}.  Quantitative viral culture was performed by standard plaque assay.  Quantitative chemokine levels were performed in 30 $\mu$L aliquots for a panel of chemokines (IL-8, MCP-1, MIP-1$\alpha$, MIP-1$\beta$, RANTES, IP-10, eotaxin) and cytokines (interferon-gamma, IL-1$\alpha$, IL-1$\beta$, IL-1RA, IL-2, IL-3, IL-4, IL-6, IL-10, IL-12p40, IL-15, IL-17, TNF$\alpha$) (Luminex Assay®, Luminex Corp.) and reported as ng/mL basal media sampled from a total volume of 4 mL.  Only IP-10, RANTES, and TNF$\alpha$ showed increases in production.  TNF activity as not incorporated in the model.  Data for other chemokines and cytokines is not shown.

@@

To provide estimates of chemokine concentrations and secretion rates present in lung tissue, chemokine levels were measured at 4-6h intervals during the first 48 h of infection in wells containing approximately one million human bronchial epithelial cells (Fig.~\ref{fig:data}).  The dynamic viral loads at these intervals in these cultures infected with seasonal H1N1 virus, pandemic H1N1 virus, and avian H5N1 virus have been reported previously \cite{Mitchell2011}.  IP-10 concentration increases were observed by 8h post-infection (p.i.), and RANTES by 16h p.i..  To estimate per-cell production rates, we extended the delay differential equation (DDE) model of Ref. \cite{Mitchell2011} to represent chemokine production from infected cells (Eq.~S1).  Model fits (Table \ref{tab:strains}) were computed for three strains using Matlab's \texttt{nlinfit} function (Levenberg-Marquardt algorithm).  The resulting chemokine production values were used in the CyCells ABM by combining IP-10 and RANTES into one aggregate rate but also reported separately (Fig.~S2).  Best-fit expression rates were similar for all strains except for significantly higher RANTES production in aH5N1.  There was no positive correlation between viral production and induced chemokine production across the three strains.

\subsection*{Stochastic modeling effects}

Unlike ordinary differential equation (ODE) and DDE models, which are deterministic, stochastic models such as CyCells can produce different results on different runs (Fig.~\ref{fig:variance}).  To test the strength of this effect, we ran each model fifty times using the default parameters given in Table 2.  $R^2$ values were calculated for individual runs versus the average of all runs.  The $R^2$ ($\pm$ one standard deviation) for avian, seasonal, and pandemic influenza are: (0.957 $\pm$ 0.150), (0.992 $\pm$ 0.013), and (0.998 $\pm$ .002) respectively (i.e. there was little variation between runs).  The three strains show different levels of replication as found in \cite{Mitchell2011}.

Overlaying multiple runs on a single plot reveals a growth rate decline at 4 days post infection, reflecting the addition of IgM.  Subsequently, for each infection the number of infected cells declines quickly after day five post infection due to the CD8 T cell response.  aH5N1 is cleared completely, sH1N1 is more slowly cleared, and pH1N1 clearance at day 6 post infection is halted as replication overtakes clearance.

The rapid production of the pH1N1 virus prevents the immune response from containing the infection (Fig.~\ref{fig:variance}C).  Because pH1N1 replicates more rapidly, the number of infected cells outstrips the ability of the T cells to find them.   Spatial effects reveal the reasons for this in simulation images and video (Fig.~\ref{fig:cycells}, Video S4).

@@

Spatial effects of viral replication and chemokine diffusion play an important role in both the spreading and clearing of infections, imaged as plaque growth over time (Fig.~\ref{fig:cycells}).  Free virus particles diffuse from virus secreting cells and infect healthy cells.  Chemokine produced by infected cells attracts T cells to the infected cells.  Although larger size of virus particles causes virus to diffuse much more slowly than chemokine molecules, chemokine decays more quickly.  These countervailing effects result in similar spatial profiles for the two particle types (Fig.~\ref{fig:cycells}). Until day 4 the plaque is dominated by active (incubating and secreting) cells, whereas there is a smaller percentage of dead cells. Over time, cells in the plaque's interior die, and active cells form a decreasing proportion of the plaque. T cells arrive at day 5 and begin killing the virus-secreting cells. By day 6 many expressing cells have been eliminated and the plaque is dominated by dead cells.  Continuous change in plaque size and T cell number is demonstrated for seasonal H1N1 virus (Videos S1-S3).

In the aH5N1 infection, the plaque is dense, allowing T cells to find secreting cells easily, and infection is eliminated.  However, secreting cells were not eliminated in either sH1N1 or pH1N1 simulations (Fig.~\ref{fig:variance}).  Secreting cells accounted for at most 10\% of the active cell population and less than 1\% of the total plaque at 6 days p.i. (Fig.~\ref{fig:plaquesize}).  T cells still accumulate, but they arrive at a lower rate than the number of infected cells that are added to the plaque.  Further, the regions of concentrated chemokine diffuse slowly, causing them to be located behind the expanding infected cell and virus locations.  It takes time for newly secreting cells to produce chemokine while pockets of high chemokine density are slow to decay.  Thus, T cells can fail to locate new regions of infected cells within the plaque.  Taken together, the delayed response and low proportion of virus secreting cells prevent T cells from completely clearing the infection.  Furthermore, increasing chemokine production fails to guide T cells to the actively secreting epithelial cells in the runaway pH1N1 infection (data not shown) suggesting that the effect of high viral production rates dominate increased chemokine production rates.

\subsection*{Temporal effects}

The notion that T cell can't find infected cells in large plaques was examined in more detail by comparing the ratios of T cells to infected cells in Fig.~\ref{fig:plaquesize}.  T cells are unable to kill virus incubating epithelial cells that have not yet presented antigen.  At day 5.5, the ratio of antigen presenting secreting epithelial cells to the total plaque size is high (Fig.~\ref{fig:cycells}-\ref{fig:plaquesize}). By day 7, this ratio is very small for both the pH1N1 and sH1N1 strains, making it much more difficult for searching T cells to find the secreting cells.  The high replication rate of pH1N1 exaggerates this effect (Fig.~\ref{fig:plaquesize}C).  Video S4 illustrates this event.

Because there is an interval between first secreted virus and the first arriving T cell, infection control cannot be completed by CD8 T cells alone.  An epithelial cell infected with pH1N1 produces new virus at the rate of 5.08e-3 PFU/s \cite{Mitchell2011}.  Assuming that secretion continues for one hour after apoptosis is initiated, the best a T cell could do is limit production to 18 new viral particles.  Thus, T cells alone cannot contain the pH1N1 infection.  In contrast, in a sH1N1 virus infection, T cells would limit a single infected cell to producting 1.4 viral particles in the one-hour window.  Avian H5N1 virus-secreting cells produce only 0.2 viral particles in the interval after induced apoptosis.  Viral clearance depends on the interaction of the rate of apoptosis, viral decay, viral production, and the viral infection probability.
